Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}. For simplicity, we will consider an infinite set of named locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.

We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  partially-ordered \textbf{carrier set} \(P\) such that every \(p \in P\) has
  finitely many predecessors equipped with a mapping \(\alpha: P \to \mc{A}\)
  quotiented under the equivalence relation \(\alpha: P \to \mc{A} \simeq
  \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\) from \(P\)
  and \(Q\), there exists a partial order isomorphism between them which
  respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
  carrier set is.
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\), which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to \mc{A}\), where \(P + Q\) is given the lexicographic ordering and \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many predecessors. They also form a monoid under parallel composition \(\alpha || \beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).

@@

We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\).

\TODO{how to do reads}

\TODO{how to do writes}

@@

\TODO{submonad rune}

\TODO{Buffer continuity}

>>>>

Another interesting use of the trace monad to build a sophisticated \isotopessa model is to build a model of TSO-style weak memory based on \citet{sparky}. For simplicity, we will consider an infinite set of named, fixed locations \(x, y, z \in \ms{Loc}\) which are subject to concurrent modification by all processors via TSO reads, writes, and fences.

We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  partially-ordered \textbf{carrier set} \(P\) such that every \(p \in P\) has
  finitely many predecessors equipped with a mapping \(\alpha: P \to \mc{A}\)
  quotiented under the equivalence relation \(\alpha: P \to \mc{A} \simeq
  \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\) from \(P\)
  and \(Q\), there exists a partial order isomorphism between them which
  respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
  carrier set is. We will write unordered pomsets using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\), which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to \mc{A}\), where \(P + Q\) is given the lexicographic ordering and \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many predecessors. They also form a monoid under parallel composition \(\alpha || \beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).

@@

We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\begin{equation}
  \dnt{\ms{read}_x} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
latter, but this definition does \textit{not} support sliding:
\begin{equation}
  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
\end{equation}
since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\). If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.

To graduate from SC concurrency to a genuine, if maximally simple, weak memory model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of lists of actions, which we will interpret as linear pomsets ordered by index.

@@

\TODO{too powerful, submonad rune of buffer continuity}
